Testing and Isolation Efficacy; Insights from a Simple Epidemic Model

### 2021-02-01 SIR_manuscript.tex
more on discussion and updated a ref in intro

Because individuals who have been tested or received a positive test are likely to self-isolate, thus reducing or eliminating  their potentially infectious contacts, the dynamics of an epidemic will depend on who gets tested and
why. For a given testing intensity (tests performed per day), testing will have a stronger effect on infection rates
the more strongly it is focused on people who are infectious for SARS-CoV-2.
This level of focus depends in turn on the purpose and design of testing programs.
(1) When testing is done for the purposes of disease surveillance \citep{foddai2020base}
we might expect tests to be assigned randomly across the population (in an ideal world: some bias in selection for testing is probably inevitable), although a stratified sampling design would be preferred for statistical efficiency.

@@

Here, we develop a compartmental model to incorporate epidemic processes and testing processes mechanistically to study the potential effect of testing strategies, testing intensity, test sensitivity and specificity, reporting time and isolation on epidemic dynamic. We investigate the effect of underlying parameters of the model on the basic reproduction number, $\Rnum$. While targeted testing strategies are always more effective than random testing, as expected, we find that in some cases the direct effect of testing is viral spread is expected to be stronger for a slow test than for a fast test. This counter-intuitive effect can occur when people are cautious when awaiting a test result, and may not be robust to second-order effects of fast testing (such as better contact tracing).

{\bf \percap testing intensity, $\rho$};
Prop. \ref{p1:rho} indicates that increasing testing intensity reduces $\Rnum$. This is sensible since as people are moved to test compartments, namely $I_p$ and $I_n$ in particular, may partially or fully self-isolate, and individuals who have tested positive, namely $I_c$ compartment, are highly likely to self-isolate. The higher probability of being subject to isolation, the lower the transmission \eqref{Lambda} will become and the lower $\Rnum$ will become. Our simulation supports this fact. However, it appeared to be hard to conclude this by directly differentiating the expression of $\Rnum$ due to its complexity.

{\bf Delay in test reporting rate, $\omega$};
Prop.~\ref{p1:omega} is the most surprising. It states that returning test results more rapidly (i.e., increasing $\omega$) does not necessarily lower $\Rnum$; whether increasing $\omega$ lowers $\Rnum$ depends on the precise combination of model parameters (test reporting rate, testing processes/strategies ($W_*$), testing efficacy ($P_I$) and isolation ($\eta_*$). Specifically, the $\partial{R_0}/\partial{\omega}>0$ suggests that reproduction number may increase as the test reporting process become faster. This can be seen from expression \eqref{Rom} in the case of perfect isolation, when $\eta_w=0$ and $\eta_c=0$. The underlying mechanism for this result is that individuals may take more precautionary behaviors (e.g., physical distancing) when they are awaiting test results compared to when they are untested or their test has returned negative. An example of a situation that would favour slow return of tests is if the test being employed produces many false negatives, because many infected individuals will believe they are negative and thus may unknowingly spread the virus to many others.  For an example of the behaviour of $\Rnum$ see Figure \ref{pan} at varying values of $\omega$. This property also provides a tool to quantify the amount of delay required in the test reporting process as a strategy to reduce $\Rnum$ and consequently control the epidemic.

{\bf The correspondence of testing strategies and \percap testing weights}; 
-All of the accessible testing data is being used for one of the other three purposes rather than surveillance. These are targeted testing strategies with a different bias.
- when people are tested for "screening", the individuals with potential higher mobility, eg. people who are getting on flights, get more tested and thus the coresponding heavier testing weight is assigned than people awaiting for a surgery and are probably going to stay in a long-term care facility and consequently less mobile and more isolated to begin with.

>>>>

Because individuals who have been tested or received a positive test are likely to self-isolate, thus reducing or eliminating  their potentially infectious contacts, the dynamics of an epidemic will depend on who gets tested and
why. For a given testing intensity (tests performed per day), testing will have a stronger effect on infection rates
the more strongly it is focused on people who are infectious for SARS-CoV-2.
This level of focus depends in turn on the purpose and design of testing programs.
(1) When testing is done for the purposes of disease surveillance \citep{foddai2020base}
we might expect tests to be assigned randomly across the population (in an ideal world: some bias in selection for testing is probably inevitable), although a stratified sampling design would be preferred for statistical efficiency \citep{graubard1996modelling} \ali{better ref?}.

@@

Mathematical modeling of infectious disease outbreaks provides insights on how testing processes influence the epidemiological processes through isolation. 
Here, we develop a compartmental SIR-type model to study the potential effect of testing strategies, testing intensity, test sensitivity and specificity, test reporting time and isolation on epidemic dynamics. 
While targeted testing strategies are always more effective than random testing, as expected, we find that in some cases the direct effect of testing is viral spread is expected to be stronger for a slow test than for a fast test. This counter-intuitive effect can occur when people are cautious when awaiting a test result, and may not be robust to second-order effects of fast testing (such as better contact tracing).

We compared two testing strategies, random and \emph{tergeted}, by incorporating the compartment-specific relative testing weights $W_S$, $W_I$ and $W_R$. Although, in this study for the purpose of simplicity and illustraition, we assumed $W_I/W_S=3$ for the targeted strategy scenario, we have not specified a particular relative testing weights corresponding to a particular targeted testing scenario. 
Modeling different targeted testing strategies, thus test-specific testing weights, requires prior information of the conditional probabilities of getting tested for people in a given compartment. This part needs to be developed further in future work. 
This can be implied when we would like to quantify and compare the effect of different levels of test forcus for infectious people on the basic reproduction number $\Rnum$, and conclude about the disease spread management. For example, when people are tested for "screening", the individuals with potential higher mobility, eg. people who are getting on flights, get more tested and thus the coresponding heavier testing weight is assigned than people awaiting for a surgery and are probably going to stay in a long-term care facility and consequently less mobile and more isolated to begin with. With our model, we would be able to compare the sensitivity of the disease dynamics, through $\Rnum$, with respect to testing processes.

{\bf \percap testing intensity, $\rho$};
Proposition \ref{prop1} part \ref{p1:rho} indicates that increasing testing intensity $\rho$ reduces $\Rnum$. This is sensible since as people are moved to test compartments, namely $I_p$ and $I_n$ in particular, may partially or fully self-isolate, and individuals who have tested positive, namely $I_c$ compartment, are highly likely to self-isolate. The higher probability of being subject to isolation, the lower the transmission \eqref{Lambda} will become and the lower $\Rnum$ will become. Our simulation supports this fact. However, it appeared to be hard to conclude this by directly differentiating the expression of $\Rnum$ due to its complexity.

{\bf Delay in test reporting rate, $\omega$};
Proposition \ref{prop1} part \ref{p1:omega} is the most interesting result. It states that returning test results more rapidly, i.e., increasing $\omega$, does not necessarily lower $\Rnum$; whether increasing $\omega$ lowers $\Rnum$ depends on the precise combination of model parameters (test reporting rate, testing processes/strategies ($W_*$), testing efficacy ($P_I$) and isolation ($\eta_*$). Specifically, the $\partial{R_0}/\partial{\omega}>0$ suggests that reproduction number may increase as the test reporting process become faster. This can be seen from expression \eqref{Rom} in the case of perfect isolation, when $\eta_w=0$ and $\eta_c=0$. The underlying mechanism for this result is that individuals may take more precautionary behaviors (e.g., physical distancing) when they are awaiting test results compared to when they are untested or their test has returned negative. An example of a situation that would favour slow return of tests is if the test being employed produces many false negatives, because many infected individuals will believe they are negative and thus may unknowingly spread the virus to many others.  For an example of the behaviour of $\Rnum$ see Figure \ref{pan} at varying values of $\omega$. This property also provides a tool to quantify the amount of delay required in the test reporting process as a strategy to reduce $\Rnum$ and consequently control the epidemic.
